Sherlock topic model paper

### 2018-05-18 main.tex
finished event segmentation results section

\subsubsection{A model to capture the moment-by-moment flow of information in naturalistic stimuli}
We fit a topic model to overlapping windows of manually annotated text descriptions of scenes from an episode of the BBC series, Sherlock. The text description contained details of the scene such as the characters, location, and a short summary of the scene (see \ref{fig:schematic} for example text, see \ref{sec:methods} for analysis details). We then transformed the text descriptions with the (same) topic model, resulting in a scene (1000) by topics (100) matrix where each row of the matrix represents a probabalistic mixture of topics discovered in that scene. We expanded this matrix from 1000 to 1976 timepoints by copying the vectors for scenes that spanned multiple TRs. As depicted in \ref{fig:model}a, the topics are sparse and change slowly over time. This gradual change was anticipated, since we used overlapping windows of text (50 sentences) as input to the model. The sparsity was also expected since the topic modeling approach we used (Latent Dirichlet Allocation) assumes a sparse Dirichlet prior. By eye, there are clear transitions from one topic `state' to the next, likely indexing scene transitions in the stimulus.  To get a better handle on this temporal structure, we computed the timepoint-by-timepoint correlation matrix of the video model (\ref{fig:model}b).  This correlation matrix reveals that the model has a strong, blocky structure along the diagonal.  Another interesting feature is that there is very little correlation between events (i.e. the off-diagonal values are small). This is an important feature of the model because the `event' representations are unique and highly discriminable.

@@

\subsubsection{Segmenting the video and recall models into events}
One prominent feature of the video and recall correlation matrices (see \ref{fig:model}) is the strong blocky structure along the diagonal.  We hypothesized that this structure arises from transient stability in the language used to describe each scene of the video. If that were true, we reasoned that it would be possible to decode which particular scene a participant is describing by comparing the average topic model for a particular recall `event' with the average topic vector for each video event. The video event that is most similar to the given recall event is the event the participant is most likely describing.  To test this idea it is first necessary to systematically parse the matrices into events.  To do this, we leveraged the event segmentation approach developed by Baldassano et al 2017.  Briefly, this algorithm implements a variant of a Hidden Markov Model to estimate k events given a (multivariate) time series and a k value (number of states). We automatically determined a k value for the video model and each participant's recall model using the following equation:

@@

Next, we tested the hypothesis that the individual `blocks' along the diagonals of the recall matrices represent the recall of a particular video event. To this end, we created a video event model, which averaged together neighboring topic vectors that the were classified to be in the same event (\ref{fig:eventseg}c), resulting in an events (34) by topics (100) matrix.  We performed the same procedure for the recall matrices (see \ref{fig:eventseg}d for example). Then, we computed the correlation between these two event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (see \ref{fig:eventseg}e). These matrices represent the extent to which each recall event is similar to each video event (for each participant). To determine which video events the recall events most likely referred to, for event recall event, we computed video event with the highest correlation (i.e. the argmax).  This is depicted in \ref{fig:eventseg}e as the cell highlighted with the black box for each column in the video-recall event correlation matrix.

Figure X displays example correlation matrices for the participants with the lowest and highest k value.  We found that the estimated k value strongly correlated with the number of (hand-annotated) correctly remembered scenes across participants (Figure Xx, r=.67, p=.003).

>>>>

\subsubsection{A model to capture the moment-by-moment flow of information in naturalistic stimuli}
We fit a topic model to overlapping windows of manually annotated text descriptions of scenes from an episode of the BBC series, Sherlock. The text description contained details of the scene such as the characters, location, and a short summary of the scene (see \ref{fig:schematic} for example text, see \ref{sec:methods} for analysis details). We then transformed the text descriptions with the (same) topic model, resulting in a scenes (1000) by topics (100) matrix where each row of the matrix represents a probabalistic mixture of topics discovered in that scene. We expanded this matrix from 1000 to 1976 timepoints by copying the vectors for scenes that spanned multiple TRs. As depicted in \ref{fig:model}a, the topics are sparse and change slowly over time. This gradual change was anticipated, since we used overlapping windows of text (50 sentences) as input to the model. The sparsity was also expected since the topic modeling approach we used (Latent Dirichlet Allocation) assumes a sparse Dirichlet prior. By eye, there are clear transitions from one topic `state' to the next, likely indexing scene transitions in the stimulus.  To get a better handle on this temporal structure, we computed the timepoint-by-timepoint correlation matrix of the video model (\ref{fig:model}b).  This correlation matrix reveals that the model has a strong, blocky structure along the diagonal.  Another interesting feature is that there is very little correlation between events (i.e. the off-diagonal values are small). This is an important feature of the model because the `event' representations are unique and highly discriminable.

@@

\subsubsection{Segmenting the video and recall models into events}
One prominent feature of the video and recall correlation matrices (see \ref{fig:model}) is the strong, blocky structure along the diagonal.  We hypothesized that this structure arises from transient stability in the language used to describe each event of the video. If that were true, we reasoned that it would be possible to decode which particular event a participant is describing by comparing the average topic model for a particular recall `event' with the average topic vector for each video event. The video event that is most similar to the given recall event is the event the participant is most likely describing.  To test this idea it is first necessary to systematically parse the matrices into events.  To do this, we leveraged the event segmentation approach developed by Baldassano et al 2017.  Briefly, this algorithm implements a variant of a Hidden Markov Model to estimate k events given a (multivariate) time series and a k value (number of states). We automatically determined a k value for the video model and each participant's recall model using the following equation:

@@

Next, we tested the hypothesis that the individual `blocks' along the diagonals of the recall matrices represent the recall of a particular video event. To this end, first we created a video event model, which was done by averaging together neighboring topic vectors that the were classified to be in the same event (\ref{fig:eventseg}c), resulting in an events (34) by topics (100) matrix.  We performed the same procedure for the recall matrices (see \ref{fig:eventseg}d for example). Then, we computed the correlation between these two event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (see \ref{fig:eventseg}e). These matrices represent the extent to which each recall event is similar to each video event (for each participant). To determine which video events the recall events most likely referred to, for each recall event, we found the index of the video event with the highest correlation (i.e. the argmax).  This is depicted in \ref{fig:eventseg}e as the cell highlighted with the black box for each column in the video-recall event correlation matrix. Notably, our algorithm suggests that the example participant recalled the events largely in order, occasionally skipping over an event.

Next, we computed a group average recall event model and video-recall event correlation matrix (\ref{fig:eventseg}f).  To do this, for each participant (and each recall event), we grouped the recall event vectors by the video event that they were most highly correlated to. We then averaged the recall event vectors across subjects. This gave an average recall event vector for all but one (of 34) events since no subjects remembered this events (according to our model). Lastly, we computed an average recall event (34) by video event (34) correlation matrix, and highlighted the highest correlation in the column with a black box (\ref{fig:eventseg}f). Notably, this matrix displayed high correlation values along the diagonal and low correlations in the off-diagonal cells.

To better visualize the relationship between the video and recall models, we embedded the video and average recall model into a 2D space (using the UMAP dimensionality reduction algorithm, for details see \ref{sec:methods}) where the points represent video/recall events and the distance between them represents their similarity in 'topic space' (\ref{fig:eventseg}g). We computed the group average transition probabilites from each event to each other event and plotted the values as lines where the tranparency is proportional to the probability of the transition.  By eye, it is clear that the two models have a very similar geometric structure, and the transition probabilities tend to be strongest between events that occured sequentially (or nearby) in time.

To explore individual variability (and consistency) in the recall event models, we embedded each participant's recall event model in the same 2D 'topics space' described above (\ref{fig:eventseg}h). Despite the variability in the number of events remembered, the overall shape of most of the participants' recall trajectories are similar to the video model (\ref{fig:eventseg}h), particularly the participants with a greater number of recalled events. To quantify the similarity between the video model and individual recall models, we considered a number of metrics.  First, we tested whether each participant's recall model matched the movie model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered. Then, we computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we permformed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants (p<.05; participants: 3-4, 8-13, 17) and the p-value for the rest of the participants was less than .25. Furthermore, the RMSD values were significantly correlated to memory performance across participants (Pearson's r=-.57, p=.016)  Next, we tested whether participants who recalled more events were also more precise in their recollections. For each participant, we computed the correlation between each recall event and its matching video event (only for the events which they recalled) and averaged the correlation within participant.  In line with our prediction, there was a strong correlation between memory performance and precision suggesting that participants who remembered more events also remembered them more veridically (Pearson's r=.74, p=.0006). Related to this last measure, we considered the distinctiveness of each recall event. That is, how unique each recall event is with respect to all other recall events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (compared to those with lower memory performance who might describe events in a more general way). To this end, we computed a 'distinctiveness' score for each participant, i.e. 1 - correlation between the given recall event and all other recall events for the participant.  Then, we averaged this measure across recall events.  We found that participants with higher memory performance also had higher distinctiveness scores (Pearson's r=.8, p=.0001). Finally, we tested whether participants with better memory performance were more likely to remember the events in order.  For each participant, we computed the Spearman rank correlation between the order of events that the participant recalled and the actual order of events, sampling out only events that were actually recalled.  We found that participants who recalled more events also recalled more of them in order (Pearson's r=.5, p=.04).
